---
title: "Dynamic Rhythms"
author: "Yishun Zhang"
date: "2025-04-21"
output: html_document
---
Number of people affected by power outages from 2014 to 2023

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
power14<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2014.csv")
power15<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2015.csv")
power16<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2016.csv")
power17<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2017.csv")
power18<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2018.csv")
power19<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2019.csv")
power20<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2020.csv")
power21<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2021.csv")
power22<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2022.csv")
power23<- read.csv("C:/Users/17756/Downloads/eaglei_outages_2023.csv")

```

Integrate the data from 2014 to 2023

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(fasttime)

dfs <- mget(paste0("power", 14:23))

dfs2 <- lapply(dfs, function(df) {
  df %>%
    mutate(
      run_start_time = as.character(run_start_time),
      run_start_time = fasttime::fastPOSIXct(run_start_time, tz = "UTC")
    )
})

df_all <- bind_rows(dfs2) %>%
  arrange(run_start_time)

head(df_all)

```

Identify the state with the most power outages in the past decade
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(maps)
library(viridis)

start_time <- as.POSIXct("2014-11-01 00:00:00", tz = "UTC")
end_time   <- as.POSIXct("2023-12-31 23:45:00", tz = "UTC")

agg <- df_all %>%
  filter(run_start_time >= start_time,
         run_start_time <= end_time) %>%
  group_by(state) %>%
  summarise(total_out = sum(customers_out, na.rm = TRUE)) %>%
  mutate(customers_day = total_out / 96,   
         region = tolower(state))         

us_map <- map_data("state") %>%
  mutate(region = tolower(region))

map_df <- left_join(us_map, agg, by = "region")

ggplot(map_df, aes(x = long, y = lat, group = group, fill = customers_day)) +
  geom_polygon(color = "white", linewidth = 0.2) +
  coord_fixed(1.3) +
  scale_fill_viridis(na.value = "grey90",
                     name = "cust·day",
                     option = "C") +
  labs(title    = "U.S. Power Outages by State, Jan–Dec 2023 (customers·day)",
       subtitle = paste(format(start_time, "%Y-%m-%d"),
                        "to",
                        format(end_time,   "%Y-%m-%d"))) +
  theme_void() +
  theme(legend.position = "right")

```

Plot the number of power outages in Florida over the past decade on the chart.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)

df_fla <- df_all %>%
  filter(state == "Florida",
         year(run_start_time) >= 2014,
         year(run_start_time) <= 2023) %>%
  mutate(
    Year = factor(year(run_start_time)),       

    date_norm = as.Date(yday(run_start_time) - 1,
                        origin = "2020-01-01")  
  )
Sys.setlocale("LC_TIME", "C")
ggplot(df_fla, aes(x = date_norm,
                   y = customers_out,
                   color = Year,
                   group = Year)) +
  geom_line(alpha = 0.8) +
  scale_x_date(date_breaks = "2 months",
               date_labels = "%b") +
  labs(
    title = "Total Customers Out in Florida by Calendar Day (2014–2023)",
    subtitle = "Each line shows one year's outages, aligned Jan–Dec",
    x = NULL,
    y = "Customers Out",
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, vjust = 0.5),
    legend.position = "right"
  )


```

The number of power outages in these four states over the past decade
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_yearly <- df_all %>%
  filter(state %in% c("Florida","California","Texas","Michigan")) %>%  
  mutate(Year = year(run_start_time)) %>%                            
  filter(Year >= 2014, Year <= 2023) %>%                             
  group_by(state, Year) %>%
  summarise(
    customers_day = sum(customers_out, na.rm = TRUE) / 96             
  ) %>%
  ungroup()

ggplot(df_yearly, aes(x = Year, y = customers_day, color = state)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2014:2023) +
  labs(
    
    title = "Annual Power Outages (Customers·Day) in Four States (2014–2023)",
    x     = "Year",
    y     = "Customers·day",
    color = "State"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```


Storm data from 2014 to 2023
```{r, echo=FALSE, message=FALSE, warning=FALSE}
storm14 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2014_c20231116.csv")
storm15 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2015_c20240716.csv")
storm16 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2016_c20220719.csv")
storm17 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2017_c20230317.csv")
storm18 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2018_c20240716.csv")
storm19 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2019_c20240117.csv")
storm20 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2020_c20240620.csv")
storm21 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2021_c20240716.csv")
storm22 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2022_c20241121.csv")
storm23 <- read.csv("C:/Users/17756/Downloads/StormEvents_details-ftp_v1.0_d2023_c20241216.csv")
```

Consolidate the storm data from the past decade
```{r, echo=FALSE, message=FALSE, warning=FALSE}

dfs <- mget(paste0("storm", 14:23))

df_storm_all <- bind_rows(dfs)

df_storm_all

```


Type of storm
```{r, echo=FALSE, message=FALSE, warning=FALSE}
unique(df_storm_all$EVENT_TYPE)
```


The number of types of storms in Florida over the past decade
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_fla_counts <- df_storm_all %>%
  filter(STATE == "FLORIDA") %>%
  count(EVENT_TYPE, name = "n") %>%
  arrange(n)

ggplot(df_fla_counts, aes(x = reorder(EVENT_TYPE, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Number of Storms by Type in Florida",
    x     = "Storm Type",
    y     = "Count"
  ) +
  theme_minimal()
```


The number of the first six types of storms over ten years, by month.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
storm_types <- c("Thunderstorm Wind", "Heavy Rain", "Flood", 
                 "Flash Flood", "Hail", "Tropical Storm")

df_fla_monthly <- df_storm_all %>%
  filter(STATE == "FLORIDA",
         EVENT_TYPE %in% storm_types) %>%
  count(EVENT_TYPE, MONTH_NAME, name = "n") %>%
  mutate(month = factor(MONTH_NAME, levels = month.name)) %>%
  arrange(EVENT_TYPE, month)

ggplot(df_fla_monthly, aes(x = month, y = n)) +
  geom_col(fill = "steelblue") +
  facet_wrap(~ EVENT_TYPE, ncol = 3, scales = "free_y") +
  scale_x_discrete(drop = FALSE) +    
  labs(
    title = "Monthly Counts of Six Storm Types in Florida (2014–2023)",
    x     = "Month",
    y     = "Number of Events"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text  = element_text(face = "bold")
  )

```


The duration of each storm.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_storm_all2 <- df_storm_all %>%
  mutate(
    begin_str = paste0(
      substr(BEGIN_YEARMONTH, 1, 4), "-",         
      substr(BEGIN_YEARMONTH, 5, 6), "-",           
      sprintf("%02d", BEGIN_DAY), " ",        
      sprintf("%04d", BEGIN_TIME)               
    ),
    end_str = paste0(
      substr(END_YEARMONTH, 1, 4), "-",
      substr(END_YEARMONTH, 5, 6), "-",
      sprintf("%02d", END_DAY), " ",
      sprintf("%04d", END_TIME)
    ),
    begin_dt = ymd_hm(begin_str, tz = "UTC"),
    end_dt   = ymd_hm(end_str,   tz = "UTC")
  ) %>%
  mutate(
    duration_mins  = as.numeric(difftime(end_dt, begin_dt, units = "mins")),
    duration_hours = as.numeric(difftime(end_dt, begin_dt, units = "hours"))
  )


head(df_storm_all2 )

```


Frequency of storms of various durations.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(df_storm_all2, aes(x = duration_hours)) +
  geom_histogram(breaks = seq(0, 100, by = 5),
                 closed = "right",
                 fill   = "skyblue",
                 color  = "grey30",
                 alpha  = 0.8) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    title = "Distribution of Event Durations (2014–2024)",
    x     = "Duration (Hours)",
    y     = "Frequency"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90")
  )


```


The number of storms of each duration.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(df_storm_all2, aes(x = EVENT_TYPE, y = duration_hours)) +
  geom_boxplot(outlier.shape = 21,
               outlier.size  = 1,
               fill           = "lightpink",
               color          = "grey40",
               alpha          = 0.7) +
  coord_flip() +  
  labs(
    title = "Event Durations by Event Type",
    x     = NULL,
    y     = "Duration (Hours)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0),
    axis.text.y = element_text(size  = 7),
    panel.grid.major.y = element_blank()
  )

```


Frequency of storms of various durations.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_fla_daily <- df_all %>% 
  filter(state == "Florida") %>% 
  mutate(date = as.Date(run_start_time))

df_fla_daily_cd <- df_fla_daily %>% 
  group_by(date) %>% 
  summarise(
    customers_day = sum(customers_out, na.rm = TRUE) / 96
  ) %>% 
  ungroup()
head(df_fla_daily_cd)

```


The number and duration of storms per day from 2014 to 2023.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)


dt <- as.data.table(df_storm_all2)[
  STATE == "FLORIDA",
  .(begin_dt = as.POSIXct(begin_dt),
    end_dt   = as.POSIXct(end_dt),
    EVENT_TYPE)
]

day_seq <- seq(as.Date("2014-11-01"), as.Date("2023-12-31"), by = "day")
dt_days <- data.table(
  day       = day_seq,
  day_start = as.POSIXct(day_seq),
  day_end   = as.POSIXct(day_seq + 1)
)

setkey(dt,      begin_dt, end_dt)
setkey(dt_days, day_start, day_end)

ov <- foverlaps(
  dt_days, dt,
  by.x    = c("day_start","day_end"),
  by.y    = c("begin_dt","end_dt"),
  type    = "any",
  nomatch = 0L
)

ov[, `:=`(
  overlap_sec = as.numeric(pmin(end_dt, day_end) - pmax(begin_dt, day_start), units = "secs"),
  event_sec   = as.numeric(end_dt - begin_dt, units = "secs")
)]
ov[, weight := overlap_sec / event_sec]

daily_count <- ov[, .(count_prop = sum(weight)), by = .(day, EVENT_TYPE)]
daily_dur   <- ov[, .(duration_hr = sum(overlap_sec) %/% 3600), by = .(day, EVENT_TYPE)]

count_wide <- dcast(daily_count, day ~ EVENT_TYPE, value.var = "count_prop", fill = 0)
setnames(count_wide,
         old = setdiff(names(count_wide), "day"),
         new = paste0("count_prop_", gsub(" ", "_", setdiff(names(count_wide), "day"))))

dur_wide <- dcast(daily_dur, day ~ EVENT_TYPE, value.var = "duration_hr", fill = 0)
setnames(dur_wide,
         old = setdiff(names(dur_wide), "day"),
         new = paste0("dur_hr_", gsub(" ", "_", setdiff(names(dur_wide), "day"))))

final_df <- data.table(day = day_seq) %>%
  merge(count_wide, by = "day", all.x = TRUE) %>%
  merge(dur_wide,   by = "day", all.x = TRUE) %>%
  (function(x){
    num_cols <- setdiff(names(x), "day")
    x[, (num_cols) := lapply(.SD, function(col) replace(col, is.na(col), 0)), .SDcols = num_cols]
    x
  })

final_df <- as_tibble(final_df)
final_df



```


The number of outages and the number of storms are combined.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_fla_daily_cd_clean <- df_fla_daily_cd %>%
  mutate(date = as.Date(date))    

final_df_clean <- final_df %>%
  rename(date = day) %>%          
  mutate(date = as.Date(date))    

merged_df <- df_fla_daily_cd_clean %>%
  inner_join(final_df_clean, by = "date") %>%
  arrange(date)

merged_df

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)

storm_cols <- names(merged_df)[3:31]

df_daily_storms <- merged_df %>%
  mutate(
    Year      = year(date),
    MonthDay  = month(date),         
    DayOfYear = yday(date)           
  ) %>%
  rowwise() %>%
  mutate(
    StormsTotal = sum(c_across(all_of(storm_cols)), na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(df_daily_storms, 
       aes(x = DayOfYear, y = StormsTotal, color = factor(Year), group = Year)) +
  geom_line(alpha = 0.8) +
  scale_x_continuous(
    breaks = yday(as.Date(paste0("2020-", 1:12, "-15"))),  
    labels = month.name
  ) +
  scale_color_viridis_d(option = "turbo", name = "Year") +
  labs(
    title    = "Total Storm Events by Calendar Day (2014–2023)",
    subtitle = "Each line is one year, aligned Jan–Dec",
    x        = NULL,
    y        = "Total Storm Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position   = "right",
    panel.grid.minor  = element_blank(),
    axis.text.x       = element_text(angle = 45, hjust = 1)
  )

```



```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(purrr)
library(tibble)
library(forcats)
library(ggplot2)

first31 <- names(merged_df)[1:31]

event_cols <- first31[
  sapply(merged_df[first31], is.numeric) &
  first31 != "customers_day"
]

good_cols <- event_cols[
  sapply(merged_df[event_cols], function(x) sd(x, na.rm = TRUE) > 0)
]

corrs <- map_dbl(good_cols, 
                 ~ cor(merged_df$customers_day, 
                       merged_df[[.x]], 
                       use = "pairwise.complete.obs"))

corr_df <- tibble(
  event       = good_cols,
  correlation = corrs
) %>%
  filter(abs(correlation) > 0.1) %>%
  arrange(correlation) %>%
  mutate(event = fct_reorder(event, correlation))

ggplot(corr_df, aes(x = event, y = correlation, fill = correlation)) +
  geom_col() +
  scale_fill_gradient2(
    low      = "steelblue", 
    mid      = "lightgray", 
    high     = "indianred",
    midpoint = 0
  ) +
  coord_flip() +
  labs(
    title    = "Correlation between First 31 Variables and Power Outages",
    subtitle = "Only events with |correlation| > 0.1",
    x        = "Variable",
    y        = "Pearson Correlation\nwith customers_day",
    fill     = "Corr"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The delay was observed from September 1 to October 1, 2017.
```{r, echo=FALSE, message=FALSE, warning=FALSE}

start <- as.Date("2017-09-01")
end   <- as.Date("2017-10-01")

win <- merged_df[merged_df$date >= start & merged_df$date <= end, ]

storm_cols <- c("count_prop_Tropical_Storm", "count_prop_Hurricane", "count_prop_Storm_Surge/Tide", "count_prop_Flood")

storm_cols_col <- c("dodgerblue", "orange", "forestgreen", "purple")
storm_cols_lty <- c(1, 2, 3, 4)

plot(win$date, win[[storm_cols[1]]],
     type  = "n",
     ylim  = c(0, max(win[, storm_cols], na.rm = TRUE)),
     xlab  = "Date",
     ylab  = "Event Counts",
     xaxt  = "n")   

axis.Date(1, at = seq(start, end, by = "5 days"), format = "%m-%d")

for (i in seq_along(storm_cols)) {
  lines(win$date, win[[storm_cols[i]]],
        col  = storm_cols_col[i],
        lty  = storm_cols_lty[i],
        lwd  = 2)
}

par(new = TRUE) 
plot(win$date, win$customers_day,
     type  = "l",
     col   = "red",
     lwd   = 2,
     axes  = FALSE,
     xlab  = "",
     ylab  = "",
     ylim  = c(0, max(win$customers_day, na.rm = TRUE)))

axis(side = 4)
mtext("Customers·day", side = 4, line = 3, col = "red")

legend("topleft",
       legend = c(storm_cols, "Customers·day"),
       col    = c(storm_cols_col, "red"),
       lty    = c(storm_cols_lty, 1),
       lwd    = 2,
       bg     = "white",
       inset  = 0.02,
       cex    = 0.8)

title("Time Histories of Event Counts and Power Outages\nFlorida 2017-09-01 to 2017-10-01")

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(purrr)
library(tidyr)
library(ggplot2)

event_cols <- c("count_prop_Tropical_Storm", "count_prop_Hurricane", "count_prop_Storm_Surge/Tide", "count_prop_Flood")
max_lag    <- 15

n <- nrow(merged_df)

ccf_df <- expand_grid(
  event = event_cols,
  lag   = 0:max_lag
) %>%
  mutate(
    corr = map2_dbl(event, lag, ~ {
      x_full <- merged_df[[.x]]
      y_full <- merged_df$customers_day
      d <- .y
      if (d == 0) {
        cor(x_full, y_full, use = "pairwise.complete.obs")
      } else {
        x <- x_full[1:(n - d)]
        y <- y_full[(d + 1):n]
        cor(x, y, use = "pairwise.complete.obs")
      }
    })
  )

ggplot(ccf_df, aes(x = lag, y = corr, color = event)) +
  geom_line(size = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = seq(0, max_lag, by = 5)) +
  labs(
    title    = "Lagged Correlation: Storms Leading Power Outages",
    subtitle = "Correlation at delay = 0…15 days",
    x        = "Delay (days before outage)",
    y        = "Pearson Correlation",
    color    = "Event Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    axis.text.x     = element_text(angle = 0)
  )

```



```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

df_model <- merged_df %>%
  arrange(date) %>%                      
  mutate(
    storm_count  = rowSums(across(starts_with("count_prop_"))),
    storm_dur_hr = rowSums(across(starts_with("dur_hr_")))
  ) %>%
  mutate(across(
    .cols = c(storm_count, storm_dur_hr),
    .fns  = list(
      lag0 = ~ lag(., 0),
      lag1 = ~ lag(., 1),
      lag2 = ~ lag(., 2),
      lag3 = ~ lag(., 3),
      lag4 = ~ lag(., 4),
      lag5 = ~ lag(., 5),
      lag6 = ~ lag(., 6),
      lag7 = ~ lag(., 7)
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  filter(!is.na(storm_count_lag7), !is.na(storm_dur_hr_lag7))

h <- 365
n <- nrow(df_model)
train_df <- slice(df_model,        1:(n - h))
test_df  <- slice(df_model, (n - h + 1):n)

lag_count_cols <- paste0("storm_count_lag", 0:7)
lag_dur_cols   <- paste0("storm_dur_hr_lag", 0:7)
fml <- as.formula(
  paste("customers_day ~", paste(c(lag_count_cols, lag_dur_cols), collapse = " + "))
)


```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(mgcv)

f_gam <- as.formula(
  paste("customers_day ~",
        paste0("s(storm_count_lag", 0:7, ")", collapse = " + "),
        "+",
        paste0("s(storm_dur_hr_lag", 0:7, ")", collapse = " + "))
)

gam_mod <- gam(f_gam, data = train_df, family = gaussian())
summary(gam_mod)
plot(gam_mod, pages=2, rug=FALSE)


```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)

test_df <- test_df %>%
  mutate(
    pred_raw  = predict(gam_mod, newdata = .),
    predicted = pmax(0, pred_raw)          
  )

MAE  <- mean(abs(test_df$predicted - test_df$customers_day))
RMSE <- sqrt(mean((test_df$predicted - test_df$customers_day)^2))
cat("Test MAE =", MAE, "\nTest RMSE =", RMSE, "\n")
Sys.setlocale("LC_TIME", "C")
ggplot(test_df, aes(x = date)) +
  geom_line(aes(y = customers_day, color = "Actual")) +
  geom_line(aes(y = predicted,     color = "Predicted")) +
  labs(title = "GAM: Test Set Predicted vs Actual",
       x = NULL, y = "Affected Customers") +
  scale_color_manual("", values = c("Actual"="black","Predicted"="blue")) +
  theme_minimal()

```